---
title: "QC and DESeq2"
subtitle: "Prime-seq analysis"
author: "Data analysis: Paulo Jannig | Karolinska Institutet"
date: "`r paste('Last update:', format(Sys.time(), '%B %d, %Y'))`"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 3
    number_sections: true
    theme: spacelab
    highlight: default
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r}
print(setwd(dirname(rstudioapi::getSourceEditorContext()$path)))
```


```{r}
# define variables ==========================================
experiment_ID <- "PJ101_TEMPLATE"
filename_prefix <- "" # leave empty if not necessary
species <- "mouse" # mouse, human, pig, rat
ensembl_version <- 115 # define ensembl release version. We've been using:
                        # 102 for human
                        # 107 for mouse GRCm38
                        # 115 for mouse GRCm39
            
file_formats <- list(
  #"png",
  "pdf"
)

# load libraries
required_Packages_Install <- c(
  "tidyverse",
  "DESeq2",
  "ggpubr",
  "RColorBrewer",
  "ggrepel",
  "ggtext",
  "biomaRt",
  "scales",
  "writexl",
  "readxl",
  "factoextra",
  "limma"
)

for(Package in required_Packages_Install){
  if(!require(Package,character.only = TRUE)) { 
    BiocManager::install(Package, dependencies=TRUE)
  }
  library(Package,character.only = TRUE)
}

# source custom functions
source("load_functions.R") # source custom functions
```

```{bash}
mkdir -p \
  ../02.results/deseq2/ \
  ../03.figures \
  ../04.supplements
```

# Import data
```{r}
counts <- readRDS(paste0("../02.results/zUMIs/",experiment_ID,".dgecounts.rds"))

metadata <- read_excel("../01.metadata/sampleInfo.xlsx")
#metadata <- read.csv("../01.metadata/sampleinfo.txt", sep = "\t", header = TRUE)

#gene.info <- read.csv(paste0("../02.results/zUMIs/",experiment_ID,".gene_names.txt"), sep = "\t")

kept_barcodes_binned <- read.table(paste0("../00.reports/zUMIs/",experiment_ID,"kept_barcodes_binned.txt"),
                             sep = ",", header = T) %>%
  dplyr::rename(barcode = XC,
                Reads = n)

```



# Metadata
```{r}
colnames(metadata)
```

```{r}
colnames(metadata)
unique(metadata$Group)
# unique(metadata$Sex)
```

If needed, use this to remove samples/groups
Removing:
```{r}
# metadata <- metadata %>%
#   dplyr::filter(Sex != "Male")
```


Arrange metadata
```{r}
metadata <- metadata %>%
  mutate(
    Group = factor(Group, levels = c(
      "Control",
      "STZ"
      ))
  ) %>%
  arrange(Group)

levels(metadata$Group)
```


# QC sequencing depth
Remove samples with low number of sequencing reads

Check number of samples:
```{r}
unique(metadata$sample)
length(unique(metadata$sample))
```

define color palette:
```{r}
palette_groups <- brewer.pal(n = 8, name = "Dark2")
# palette_groups <- brewer.pal(n = 12, name = "Paired")
# palette_groups <- c("black", "#1F78B4")
```

## Number of reads per sample
```{r}
readspersample <- left_join(metadata, kept_barcodes_binned, by = "barcode") %>%
  mutate(sample = fct_reorder(sample, Reads, .desc = TRUE))
```

```{r}
y_max <- max(readspersample$Reads)
y_min <- min(readspersample$Reads)
y_lim <- y_max * 1.1
sample_number <- length(readspersample$sample)

# Create the bar plot
ggplot(readspersample, aes(x = sample, y = Reads, fill = Group)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = y_max, linetype = "dashed", size = 0.25) +
  geom_hline(yintercept = y_min, linetype = "dashed", size = 0.25) +
  annotate("text", x = 1, y = y_max, label = paste("Max:", scales::comma(y_max)),
           vjust = -0.3, hjust = 0, size = 6 / ggplot2:::.pt) +
  annotate("text", x = sample_number, y = y_min, label = paste("Min:", scales::comma(y_min)), 
           vjust = 1.5, hjust = 1, size = 6 / ggplot2:::.pt) +
  labs(x = "Sample", y = "Reads", title = "Reads per Sample by Group") +
  ggplot2::scale_fill_manual(values = palette_groups) +
  gg_theme() + 
  theme(
    axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1),
    legend.position = "top"
  )  + 
  coord_cartesian(xlim = c(0, sample_number+1), ylim = c(0, y_lim), expand = FALSE)
```

Save plot
```{r}
filename <- "QC_Reads_per_sample"

for (file_format in file_formats) {
  ggsave_fixed(
    plot = ggplot2::last_plot() + gg_theme() + 
      theme(axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1),
            legend.position = "top"
            ), 
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = sample_number*3,
    plot_height = 25,
    width = sample_number*5,
    height = 60
  )
}
```


If needed, use this to remove samples
```{r}
metadata <- metadata %>%
  #dplyr::filter(Sample != "F5_1") %>%
  #dplyr::filter(Sample != "")
  filter(!(sample %in% c("M2-2", 
                         "M1-3")))

```

# Make inex matrix (Counts data frame)
```{r}
frds_inex <- as.matrix(counts$umicount$inex$all)
# rownames(frds_inex) <- factor(rownames(frds_inex),
#                                 levels = gene.info$gene_id,
#                                 labels = gene.info$gene_id)
frds_inex.sum <- aggregate(frds_inex, 
                             list(row.names(frds_inex)), 
                             sum)
rownames(frds_inex.sum) <- frds_inex.sum$Group.1
frds_inex.sum$Group.1 <- NULL

counts.df <- frds_inex.sum %>%
  dplyr::select(metadata$barcode) %>%
  rename_all(~metadata$sample)
```

Check if there are duplicated ensembl_gene_id
```{r}
duplicated_rows <- data.frame(rownames(counts.df)[duplicated(str_remove(rownames(counts.df), "\\..*"))])
duplicated_rows
```

Removing genes ending with _PAR_Y
```{r}
counts.df <- counts.df[!grepl("_PAR_Y", rownames(counts.df)), ]
```

Remove gene version
```{r}
rownames(counts.df) <- str_remove(rownames(counts.df), "\\..*")
head(counts.df)
```


# DESeq2
make sure columns of the count matrix and the rows of the column data (information about samples) are in the same order
```{r}
colnames(counts.df)
metadata$sample
colnames(counts.df)==metadata$sample

metadata$barcode
```

returns the names of all columns in metadata that are factors.
```{r}
names(metadata)[sapply(metadata, is.factor)]
```


```{r}
# DEseq2
dds <- DESeqDataSetFromMatrix(
  countData = counts.df,
  colData = metadata,
  design = ~ Group
  # design = ~ Sex + Group + Sex:Group
  # design = ~ Donor + Treatment + Time
  # design = ~ Donor + Treatment + Time + Treatment:Time


)
## calculate size factor
dds <- estimateSizeFactors(dds)
```

Check how many genes in total and levels
```{r}
nrow(dds)
levels(dds$Group)
```

Count how many genes with 0 counts (TRUE)
```{r}
table(rowSums(assay(dds, "counts")) == 0)
```

## biomaRt annotation
retrieve gene ensembl annotation 

```{r}
# If biomaRt fails, check:
# https://stackoverflow.com/questions/77370659/error-failed-to-collect-lazy-table-caused-by-error-in-db-collect-using

# # Run
# install.packages("devtools")
#devtools::install_version("dbplyr", version = "2.3.4")

# Retrieve the full list of ensembl_gene_id with attributes of interest
annotation <- biomaRt::getBM(
  attributes = c(
    "ensembl_gene_id", # Get the Ensembl gene ID
    symbol, # Gene symbol
    #"external_gene_name", # Gene symbol
    "chromosome_name", # Chromosome number
    "gene_biotype" # Gene biotype (protein-coding, lncRNA, etc.)
  ),
  mart = useEnsembl(
    biomart = "ensembl", # Use the "ensembl" BioMart database
    mirror = "useast",
    dataset = dataset, # Specify the dataset
    version = ensembl_version # Specify the Ensembl version
  )
)

mito_genes <- annotation %>%
  filter(chromosome_name == "MT")

rRNA_genes <- annotation %>%
  filter(str_detect(gene_biotype, "^rRNA"))

RiboRelated_genes <- annotation %>%
  filter(str_detect(!!sym(symbol), "(?i)^RPL|^RPS"))

```

# QC DESeq2

## QC summary table
Consider adjusting for number of factors (e.g. Group × Time) and using facet_wrap(~Factor) for clearer visualization
```{r}
qc.tb <- data.frame(
  Sample = metadata$sample,
  gene_number = colSums(counts(dds) > 0),
  MT_proportion = colSums(counts(dds)[rownames(dds) %in% mito_genes$ensembl_gene_id, ]) / colSums(counts(dds)) * 100,
  rRNA_proportion = colSums(counts(dds)[rownames(dds) %in% rRNA_genes$ensembl_gene_id, ]) / colSums(counts(dds)) * 100,
  Rb_proportion = colSums(counts(dds)[rownames(dds) %in% RiboRelated_genes$ensembl_gene_id, ]) / colSums(counts(dds)) * 100,
  Group = factor(metadata$Group, levels = levels(metadata$Group))
)

```

```{r}
# Define a plotting function
plot_qc_metric <- function(data, 
                           metric, 
                           y_label) {

  y_values <- data[[metric]]
  y_max <- max(y_values, na.rm = TRUE)
  y_min <- min(y_values, na.rm = TRUE)
  y_lim <- ceiling(y_max * 1.1)
  sample_number <- nrow(data)

  ggplot(data, aes(x = Sample, y = .data[[metric]], fill = Group)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = y_max, linetype = "dashed", size = 0.25) +
    geom_hline(yintercept = y_min, linetype = "dashed", size = 0.25) +
    annotate("text",
      x = 1, y = y_max,
      label = paste("Max:", scales::comma(y_max, decimal.mark = ".", big.mark = "")),
      vjust = -0.3, hjust = 0, size = 6 / ggplot2:::.pt
    ) +
    annotate("text",
      x = sample_number, y = y_min,
      label = paste("Min:", scales::comma(y_min, decimal.mark = ".", big.mark = "")),
      vjust = 1.5, hjust = 1, size = 6 / ggplot2:::.pt
    ) +
    labs(x = "Sample", y = y_label, title = y_label) +
    #scale_fill_brewer(palette = palette, name = "Group") +
    ggplot2::scale_fill_manual(values = palette_groups) +
    gg_theme() +
    theme(
      axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1),
      legend.position = "top"
    ) +
    coord_cartesian(xlim = c(0, sample_number + 1), ylim = c(0, y_lim), expand = FALSE)
}
```

```{r}
# Plot gene number
plot_gene_number <- plot_qc_metric(qc.tb, "gene_number", "Detected Genes")
# Plot MT proportion
plot_mt <- plot_qc_metric(qc.tb, "MT_proportion", "Mitochondrial %")
# Plot rRNA proportion
plot_rrna <- plot_qc_metric(qc.tb, "rRNA_proportion", "rRNA %")
# Plot Rb proportion
plot_rb <- plot_qc_metric(qc.tb, "Rb_proportion", "Ribo-related proteins %")
```


```{r}
# Arrange plots together
ggpubr::ggarrange(plot_gene_number, 
                  plot_mt, 
                  plot_rrna, 
                  plot_rb, 
                  ncol = 2, 
                  nrow = 2, 
                  common.legend = TRUE,
                  widths = 1,
                  heights = 1,
                  hjust = -1,
                  align = "hv")

```

Save plot
```{r}
filename <- "QC_plot"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = length(qc.tb$Sample)*10,
    plot_height = 90,
    width = (length(qc.tb$Sample)*10)*1.1,
    height = 100 
  )
}
```

## QC Filtering
```{r}
sprintf("original gene number: %s; original sample number: %s", nrow(dds), ncol(dds))
```
Create QC table and filter samples by detected genes % mitochondrial and rRNA content
```{r}
cutoff_gene_number <- 12000
cutoff_mt_prop <- 30
cutoff_rRNA_prop <- 15

qc.tb <- qc.tb %>%
  mutate(
    keep = 
      gene_number >= cutoff_gene_number & 
      MT_proportion <= cutoff_mt_prop & 
      rRNA_proportion <= cutoff_rRNA_prop
  )
qc.tb
```

Remove samples with low number of detected genes and/or high mtDNA/rRNA content
```{r}
dds.filter.sample <- dds[, qc.tb$keep]
setdiff(colnames(dds), colnames(dds.filter.sample))
```

Filter genes with low counts

Choose one of the options below:
```{r}
# # Option 1
# ## keep genes with non-zero normalized counts
# keep.genes <- rownames(dds.filter.sample)[
#   which(
#     rowSums(counts(dds.filter.sample, normalized=TRUE)) >= 1
#     )
# ]
# sprintf("Keep genes with non-zero normalized counts")

# Option 2
## Keep genes whose total normalized counts across all samples are at least "cutoff"
# cutoff <- 20 # set cutoff
# keep.genes <- rownames(dds.filter.sample)[
#   which(
#     rowSums(counts(dds.filter.sample, normalized=TRUE)) >= cutoff
#     )
# ]
# sprintf("Keep genes whose total normalized counts across all samples are at least %s", cutoff)

# # Option 3
# ## Keep genes where at least X samples (number_of_samples) have normalized counts ≥ "cutoff"
# number_of_samples <- 4 # set samples cutoff
# cutoff <- 10 # set counts cutoff
# keep.genes <- rownames(dds.filter.sample)[
#   which(
#     rowSums(counts(dds.filter.sample, normalized=TRUE) >= cutoff ) >= number_of_samples
#     )
# ]
# sprintf("Keep genes where at least %s samples have normalized counts ≥ %s", number_of_samples, cutoff)

# # Option 4
## Keep genes where at least X samples (smallestGroupSize) have normalized counts ≥ "cutoff" (increase threshold if necessary)
cutoff <- 1 # set cutoff

### Calculte smallestGroupSize
smallestGroupSize <- metadata %>%
  dplyr::group_by(Group) %>%
  #dplyr::group_by(group, genotype) %>%
  dplyr::summarise(samples = dplyr::n()) %>%
  pull(samples) %>%
  min()
sprintf('smallest group size: %s', smallestGroupSize)

keep.genes <- rownames(dds.filter.sample)[
  which(rowSums(counts(dds.filter.sample, normalized = TRUE) >= cutoff) >= smallestGroupSize)
]
sprintf('number of genes expressed in more than %s samples with counts greater than or equal to %s: %s', smallestGroupSize, cutoff, length(keep.genes))

```

Remove unwanted gene categories
```{r}
excluded.genes <- union_all(
  mito_genes$ensembl_gene_id, # removes mitochondrial-encoded genes
  rRNA_genes$ensembl_gene_id # remove rRNA genes
  # RiboRelated_genes$ensembl_gene_id # Optionally, remove ribosome-related genes
)
sprintf("%s genes were removed", length(excluded.genes))
```

Final set of retained genes
```{r}
retained_genes <- setdiff(keep.genes, excluded.genes)
sprintf("%s genes were retained", length(retained_genes))
```

Filter the DESeqDataSet
```{r}
dds.filter <- dds.filter.sample[retained_genes, ]
sprintf('retained gene number: %s; retained sample number: %s', nrow(dds.filter), ncol(dds.filter))
#dim(dds.filter)
```

check which samples were removed
```{r}
setdiff(dds$sample, dds.filter$sample)
```

update metadata:
```{r}
metadata <- metadata %>%
  dplyr::filter(!(sample %in% setdiff(dds$sample, dds.filter$sample)))
```

save metadata
```{r}
filename <- "metadata"
write.csv(metadata, paste0(
  "../01.metadata/",
  filename_prefix,
  filename, ".csv"
),
row.names = F, quote = F
)

saveRDS(metadata, paste0(
  "../01.metadata/",
  filename_prefix,
  filename,
  ".Rds"
))
```

## Final Metadata table
```{r}
filename <- "metadata"
metadata %>% DT::datatable(
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = nrow(metadata),
    dom = "Blfrtip",
    buttons = list(
      list(extend = "csv", filename = paste0(filename_prefix, filename)),
      list(extend = "excel", filename = paste0(filename_prefix, filename))
    )
  )
)
```

## QC summary table after filtering
```{r}

qc.tb_filtered <- data.frame(
  Sample = metadata$sample,
  gene_number = colSums(counts(dds.filter) > 0),
  MT_proportion = colSums(counts(dds.filter)[rownames(dds.filter) %in% mito_genes$ensembl_gene_id, ]) / colSums(counts(dds.filter)) * 100,
  rRNA_proportion = colSums(counts(dds.filter)[rownames(dds.filter) %in% rRNA_genes$ensembl_gene_id, ]) / colSums(counts(dds.filter)) * 100,
  Rb_proportion = colSums(counts(dds.filter)[rownames(dds.filter) %in% RiboRelated_genes$ensembl_gene_id, ]) / colSums(counts(dds.filter)) * 100,
  Group = factor(metadata$Group, levels = levels(metadata$Group))
)

```

```{r}
# Plot gene number
plot_gene_number <- plot_qc_metric(qc.tb_filtered, "gene_number", "Detected Genes")
# Plot MT proportion
plot_mt <- plot_qc_metric(qc.tb_filtered, "MT_proportion", "Mitochondrial %")
# Plot rRNA proportion
plot_rrna <- plot_qc_metric(qc.tb_filtered, "rRNA_proportion", "rRNA %")
# Plot Rb proportion
plot_rb <- plot_qc_metric(qc.tb_filtered, "Rb_proportion", "Ribo-related proteins %")
```

```{r}
# Arrange plots together
ggpubr::ggarrange(plot_gene_number, 
                  plot_mt, 
                  plot_rrna, 
                  plot_rb, 
                  ncol = 2, 
                  nrow = 2, 
                  common.legend = TRUE,
                  widths = 1,
                  heights = 1,
                  hjust = -1,
                  align = "hv")

```

Save plot
```{r}
filename <- "QC_plot_filtered"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = length(qc.tb$Sample)*10,
    plot_height = 90,
    width = (length(qc.tb$Sample)*10)*1.1,
    height = 100 
  )
}
```

# DESeq2

```{r}
design(dds.filter)

dds <- DESeq(dds.filter)

# dds <- DESeq(dds.filter, 
#              betaPrior = FALSE)
resultsNames(dds)
```

Check levels
```{r}
levels(dds$Group)
```


if necessary, change factor order:
```{r}
# dds$Group <- factor(
#   dds$Group,
#   levels = c(
#     "STZ",
#     "Control"
#   )
# )
# levels(dds$Group)

```


save dds file
```{r}
filename <- "dds"
saveRDS(dds, file = paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename, ".Rds"
))
```

## biomaRt annotation
retrieve gene ensembl annotation 

```{r}
# If biomaRt fails, check:
# https://stackoverflow.com/questions/77370659/error-failed-to-collect-lazy-table-caused-by-error-in-db-collect-using

# # Run
# install.packages("devtools")
#devtools::install_version("dbplyr", version = "2.3.4")

# Retrieve the list of attributes for genes in dds

annotation <- biomaRt::getBM(
  filters = "ensembl_gene_id",
  attributes = c(
    "ensembl_gene_id", # Get the Ensembl gene ID
    symbol, # Gene symbol
    #"external_gene_name", # Gene symbol
    "chromosome_name", # Chromosome number
    "gene_biotype" # Gene biotype (protein-coding, lncRNA, etc.)
  ),
  values = rownames(dds),
  mart = useEnsembl(
    biomart = "ensembl", # Use the "ensembl" BioMart database
    dataset = dataset, # Specify the dataset
    # mirror = "useast",
    version = ensembl_version # Specify the Ensembl version
  )
)

```


Save biomaRt annotation
```{r}
filename <- "biomart_annotation"
write.csv(annotation, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
),
row.names = F, quote = F
)

saveRDS(annotation, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".Rds"
))
```


## Normalized counts

```{r}
normalized_counts <- as.data.frame(counts(dds, normalized = TRUE))

# Join with annotation and relocate relevant columns
normalized_counts <- normalized_counts %>%
  rownames_to_column("ensembl_gene_id") %>%
  dplyr::left_join(annotation, by = "ensembl_gene_id") %>%
  relocate(all_of(c(symbol, "gene_biotype", "chromosome_name")), .after = "ensembl_gene_id")
  # relocate(symbol, gene_biotype, chromosome_name, gene_biotype, .after = ensembl_gene_id)

# Replace empty gene symbols by ensembl_gene_id
normalized_counts[, symbol] <- ifelse(normalized_counts[, symbol] == "" | is.na(normalized_counts[, symbol]), 
                                   normalized_counts$ensembl_gene_id, normalized_counts[, symbol])

# Ensure ensembl_gene_id isn't blank by setting it to symbol where needed
normalized_counts$ensembl_gene_id <- ifelse(normalized_counts$ensembl_gene_id == "", 
                                            normalized_counts[, symbol], normalized_counts$ensembl_gene_id)

# Handle NA values for ensembl_gene_id by replacing them with symbol
na_symbol <- is.na(normalized_counts$ensembl_gene_id)
normalized_counts[na_symbol, "ensembl_gene_id"] <- normalized_counts[na_symbol, symbol]

```

Save normalized counts
```{r}
filename <- "deseq2_normalized_counts"
write.csv(normalized_counts, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
),
row.names = F, quote = F
)

saveRDS(normalized_counts, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".Rds"
))
```

```{r echo=FALSE}
cat("Saved file:", paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
))
```

## log2 Normalized counts
```{r}
normalized_counts.log2 <- as.data.frame(log2(counts(dds, normalized=TRUE))) %>%
  rownames_to_column("ensembl_gene_id") %>%
  dplyr::left_join(annotation, by = "ensembl_gene_id") %>%
  relocate(all_of(c(symbol, "gene_biotype", "chromosome_name")), .after = "ensembl_gene_id")
  # relocate(symbol, gene_biotype, chromosome_name, gene_biotype, .after = ensembl_gene_id)

# replace empty gene symbols by ensembl_gene_id
normalized_counts.log2[, symbol] <- ifelse(normalized_counts.log2[, symbol] == "", normalized_counts.log2$ensembl_gene_id, normalized_counts.log2[, symbol])
normalized_counts.log2[, "ensembl_gene_id"] <- ifelse(normalized_counts.log2[, "ensembl_gene_id"] == "", 
                                normalized_counts.log2[, symbol], normalized_counts.log2[, "ensembl_gene_id"])


# replace NA gene symbols by ensembl_gene_id
na_symbol <- is.na(normalized_counts.log2[, "ensembl_gene_id"])
normalized_counts.log2[na_symbol, "ensembl_gene_id"] <- normalized_counts.log2[na_symbol, symbol]
```

Save log2 normalized counts
```{r}
filename <- "deseq2_normalized_counts_log2"
write.csv(normalized_counts.log2, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
),
row.names = F, quote = F
)

saveRDS(normalized_counts.log2, paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".Rds"
))
```

```{r echo=FALSE}
cat("Saved file:", paste0(
  "../02.results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
))
```

# Supplementary table
```{r}
filename <- "deseq2_normalized_counts"
excel <- list(
  normalized_counts = normalized_counts
)
write_xlsx(excel, paste0(
  "../04.supplements/",
  filename_prefix,
  filename,
  ".xlsx"
))

filename <- "deseq2_normalized_counts_log2"
excel <- list(
  log2_normalized_counts = normalized_counts.log2
)
write_xlsx(excel, paste0(
  "../04.supplements/",
  filename_prefix,
  filename,
  ".xlsx"
))
```

# PCA plots

## for PCA analysis
calculate size factor
```{r}
dds <- estimateSizeFactors(dds)
```

Quick PCAs
```{r}
vsd <- DESeq2::vst(dds, blind = TRUE)
DESeq2::plotPCA(vsd, intgroup = c("Group"))
# DESeq2::plotPCA(vsd, intgroup = c("Group", "Sex"))
```


```{r}
rld <- DESeq2::rlog(dds, blind = TRUE)
DESeq2::plotPCA(rld, intgroup = c("Group"))
# DESeq2::plotPCA(vsd, intgroup = c("Group", "Sex"))
```

## vsd
DESeq2::plotPCA uses ntop=500 top features by variance

```{r}
vsd <- DESeq2::vst(dds, blind = TRUE)
```

Define factors - adjust factor1 and factor2 according to the number of factors
```{r}
design(dds)
```

```{r}
factor1 <- "Group"
# factor2 <- "Sex"

```

Check how many variable within factors
```{r}
levels(dds$Group)
# levels(dds$Sex)
```

Choose color palette
```{r}
palette_groups <- brewer.pal(n = 8, name = "Dark2")
# palette_groups <- brewer.pal(n = 12, name = "Paired")
# palette_groups <- c("black", "#1F78B4")
```

% total variance explained by each principal component
```{r}
pca <- prcomp(t(assay(vsd)))
fviz_screeplot(pca, ncp = 10) + gg_theme()
```

With top 500 variable gene selection (as in DESeq2::plotPCA)
```{r}
top_var_genes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 500)
pca <- prcomp(t(assay(vsd)[top_var_genes, ]))
fviz_screeplot(pca, ncp = 10) + gg_theme()
```


```{r}
## Calculate PCA data
pca_data <- DESeq2::plotPCA(vsd, 
                            intgroup = factor1,
                            # intgroup = c(factor1,factor2),
                            returnData = TRUE,
                            ntop = 500,
                            #ntop = nrow(assay(vsd)) # use this to include all genes, not only top 500
                            )
### Calculating the percentage of variance explained by PC1 and PC2
percent_var <- round(100 * attr(pca_data, "percentVar"))

segments <- pca_data %>%
  dplyr::group_by(!!sym(factor1)) %>%
  # dplyr::group_by(!!sym(factor1), !!sym(factor2)) %>%
  dplyr::summarise(xend = mean(PC1), yend = mean(PC2))
# pca_data <- merge(pca_data, segments, by = c(factor1, factor2))
pca_data <- merge(pca_data, segments, by = factor1)
```

### with segments
```{r}
pca_plot <- pca_data %>%
  ggplot2::ggplot(ggplot2::aes(PC1, PC2)) +
    ggplot2::geom_point(
    data = segments,
    ggplot2::aes(
      x = xend, y = yend,
      color = !!sym(factor1)
    ),
    size = 0.5
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = PC1, y = PC2, xend = xend, yend = yend,
      fill = !!sym(factor1),
      color = !!sym(factor1)
    ),
    linewidth = 0.3,
    linetype = "solid"
  ) +
  ggplot2::geom_point(
    aes(
      color = !!sym(factor1),
      fill = !!sym(factor1),
      # shape = !!sym(factor2)
    ),
    size = 2,
    #shape = 21,
    # color = "black",
  ) +
  ggplot2::xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ggplot2::ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  gg_theme() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::scale_shape_manual(values = c(21:25)) +
  # ggplot2::scale_color_manual(values = viridis::viridis(3 + 1)) +
  # ggplot2::scale_fill_manual(values = viridis::viridis(3 + 1)) +
  ggplot2::scale_color_manual(values = palette_groups) +
  ggplot2::scale_fill_manual(values = palette_groups) +
  scale_x_continuous(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  ggtitle("**Principal component analysis**")
pca_plot 
```

Add sample labels
```{r}
pca_plot + geom_text_repel(aes(label = name),
  size = 3 / ggplot2:::.pt,
  nudge_x = 1,
  nudge_y = -1,
  min.segment.length = 0,
  # force = 3,
  # box.padding = unit(0.1, "lines"),
  segment.size = 0.25,
  segment.colour = "black"
)
```



Save plot
```{r}
filename <- "PCA_plot_vsd_segment"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 30,
    plot_height = 30,
    width = 50,
    height = 50
  )
}
```


### with segments + ellipse
```{r}

pca_plot + ggforce::geom_mark_ellipse(
  aes(
    fill = !!sym(factor1),
    # group = interaction(!!sym(factor1), !!sym(factor2)),  # separate ellipses when having 2 factors design
    color = !!sym(factor1)
  ),
  expand = unit(1, "mm"),
  size = 0,
  tol = 0.001
)
```

```{r}

pca_plot +
  stat_ellipse(
    data = pca_data,
    aes(x = PC1, y = PC2,
        color = !!sym(factor1),
        # group = interaction(!!sym(factor1), !!sym(factor2)),
        fill  = !!sym(factor1)),
    type = "norm", 
    level = 0.68, 
    geom = "polygon",
    alpha = 0.3, 
    color = NA, 
    #position = "identity",
    inherit.aes = TRUE
     )
```


Save plot
```{r}
filename <- "PCA_plot_vsd_segment_ellipse"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format 
    ),
    units = "mm",
    plot_width = 30,
    plot_height = 30,
    width = 50,
    height = 50
  )
}
```




### only ellipses
```{r}
pca_plot <- pca_data %>%
  ggplot2::ggplot(ggplot2::aes(PC1, PC2)) +
  # ggplot2::geom_point(size = 2)
  ggplot2::geom_point(
    aes(
      color = !!sym(factor1),
      fill = !!sym(factor1),
      # shape = !!sym(factor2)
    ),
    size = 2,
    # shape = 21,
    # color = "black",
  ) +
  ggplot2::xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ggplot2::ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  gg_theme() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::scale_shape_manual(values = c(21:25)) +
  # ggplot2::scale_color_manual(values = viridis::viridis(3 + 1)) +
  # ggplot2::scale_fill_manual(values = viridis::viridis(3 + 1)) +
  ggplot2::scale_color_manual(values = palette_groups) +
  ggplot2::scale_fill_manual(values = palette_groups) +
  ggforce::geom_mark_ellipse(
    aes(fill = !!sym(factor1), 
        # group = interaction(!!sym(factor1), !!sym(factor2)),  # separate ellipses when having 2 factors design
        color = !!sym(factor1)),
    expand = unit(1, "mm"),
    size = 0,
    tol = 0.001
  ) +
  scale_x_continuous(expand = c(0.2, 0.2)) +
  scale_y_continuous(expand = c(0.2, 0.2)) +
  ggtitle("**Principal component analysis**")
pca_plot + geom_text_repel(aes(label = name),
    size = 3 / ggplot2:::.pt,
    nudge_x = 1,
    nudge_y = -1,
    min.segment.length = 0,
    #force = 3,
    #box.padding = unit(0.1, "lines"),
    segment.size=0.25,
    segment.colour = "black"
  )

```

Save plot
```{r}
filename <- "PCA_plot_vsd_ellipse"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 30,
    plot_height = 30,
    width = 50,
    height = 50
  )
}
```


```{r}
pca <- prcomp(t(assay(vsd)[top_var_genes, ]))

# 1. Make a mapping table with unique rownames
gene_symbol <- annotation %>%
  filter(ensembl_gene_id %in% rownames(pca$rotation)) %>%
  mutate(symbol = if_else(
    is.na(!!sym(symbol)) | !!sym(symbol) == "", ensembl_gene_id, !!sym(symbol)
  )) %>%
  # Make labels unique (e.g., "Actb", "Actb_1", "Actb_2")
  mutate(symbol = make.unique(symbol)) %>%
  dplyr::select(ensembl_gene_id, symbol)

# 2. Apply the new labels safely
new_labels <- gene_symbol$symbol
names(new_labels) <- gene_symbol$ensembl_gene_id

# 3. Replace rownames
rownames(pca$rotation) <- new_labels[rownames(pca$rotation)]
```

Top 10 genes contributing most to PC1
```{r}
fviz_contrib(pca, choice = "var", axes = 1, top = 10) +
  gg_theme() +
  theme(axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1)) +
  xlab("")
```



# Sample-to-sample distances
```{r}
vsd <- vst(dds.filter, blind = TRUE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Group, 
                                    vsd$sample, sep = " - ")
colnames(sampleDistMatrix) <- NULL
heatmap <- pheatmap::pheatmap(sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  show_rownames = T, show_colnames = T,
  cellwidth = 6,
  cellheight = 6,
  treeheight_row = 10,
  treeheight_col = 10,
  fontsize = 6,
  border_color = NA,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
)
```

```{r}
# Save plot
filename <- "sample_distances"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03.figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 7 * length(metadata$sample),
    height = 4 * length(metadata$sample)
  )
}
```

# R session info
```{r}
utils:::print.sessionInfo(sessionInfo()[-8])
```

# References

- Love MI, Huber W, Anders S. 2014. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol 15:550. <a href="https://doi.org/10.1186/s13059-014-0550-8" target="_blank" rel="noopener">Link</a>

<a href="https://bioconductor.org/help/course-materials/2022/CSAMA/lab/2-tuesday/lab-03-rnaseq/rnaseqGene_CSAMA2022.html" target="_blank" rel="noopener">RNA-seq workflow: gene-level exploratory analysis and differential expression</a>

<a href="https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html" target="_blank" rel="noopener">Analyzing RNA-seq data with DESeq2</a>


